Inferring the Recombinant Genealogy of a Pandemic Pathogen in Real Time

### 2025-09-05 latex/main.tex
Fix quotes, and XBB reference

UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
UShER has generated phylogenies updated daily \cite{McBroome2021} and
has proven to be a powerful resource that successfully enabled the world
to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
These phylogenies are used to identify and monitor
potentially epidemiologically important “Pango” lineages \cite{Rambaut2020,McBroome2024},
which might have acquired features to become Variants of Concern (VOC),
for example, increased transmissibility, disease severity, and immune escape.
Beyond supporting real-time genomic epidemiology,
UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
UShER uses a tree to represent evolutionary change via mutation, assuming no recombination \cite{Turakhia2021}.
However, given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2022},
the tree-like representation of UShER is no longer tenable.

@@

Here, we present
(1) sc2ts, a new method to infer ARGs in real time by building genealogies forward in time, and
(2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences.
We used the “Viridian” dataset, which is purged of various systematic errors and artifacts \cite{Hunt2024} and
show that this ARG is highly congruent to an UShER phylogeny built from the same dataset.
Compared to previously identified recombinants,
we show that this ARG contains the vast majority of epidemiologically relevant signals of recombination,
while adding significant nuance and texture.
This ARG – which captures the major signals of evolution
through phylogenetic relationships, point mutations, deletions, and recombination –
is a resource of unprecedented detail and precision, and
should provide a strong basis for future studies of SARS-CoV-2.

@@

Sc2ts uses collection dates to build a time-resolved ARG.
When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions),
likely due to errors in their collection dates
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1,
which is estimated to have emerged in November, 2021; REF).
To preclude adding such “time traveling” samples, we filtered samples by “HMM cost”.
We define the “HMM cost” of a sample to be
the total penalty cost for switches and mismatches (mutations) in its copying path.
We parameterize the LS model such that each mismatch has a penalty cost of one
and a switch has a penalty cost of $k$ (here, set to 4),
which determines how much mutation is favored over recombination when accounting for a 
sample (Figure~\ref{fig:methodology}).
If the HMM cost of a sample exceeds a threshold (here, set to 7),
sc2ts does not immediately add the sample to the ARG.

@@

Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] “problematic sites” masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods).
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow for multiple sibling origination nodes (Table SX; Document S1).

@@

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2022}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an “X”), to which we refer as “Pango X”.
Below, we summarize our analysis of the Pango Xs in the ARG
(for extended analyses, see Document S1 and Document S2).

We find strong evidence for the recombination events leading to 17 Pango Xs in the ARG (Table 1),
which passed the detection-level sc2ts quality control filters (Figure S1).
These events are supported by
high numbers of averted mutations (Table 1) and
high numbers of loci matching their suggested parents (see their copying patterns in Figure SX).
Among these Pango Xs are
XA, which is the first designated Pango X that showed signs of community transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2022}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation (\ref{fig:subgraphs}; Table 1),
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

Furthermore, some of the Pango Xs above originate from shared recombination events in the ARG (Table 1).
An example of a set of “nested” closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent events in the ARG (\ref{fig:subgraphs}).
These Pango Xs were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
\caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated recombination node (A); the recombination node associated with a group of “nested” Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C). For more detailed visualizations of these subgraphs, see Document S2.}
\label{fig:subgraphs}
\end{figure}

@@

We also compared the sensitivity of sc2ts and rebar.
Of all the “robust” recombination events detected by sc2ts,
167 (43\%) were also detected as recombinants by rebar (STAR Methods).
We hypothesize that this disagreement arose from differences in sensitivity between these methods.
Consistent with this, we find that rebar needs a higher level of divergence between parents
to correctly classify a sequence as recombinant, as shown by
a shorter time to their MRCA (\ref{fig:averted_mutations}) and
lower distance between their Pango identity (\ref{fig:averted_mutations}).

@@

Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination, as measured by the number of resulting mutations (\ref{fig:methodology}).
The more mutations are “averted” by allowing recombination, the stronger the evidence for recombination.
We calculated the number of averted mutations for all the “robust” recombination events (STAR Methods),
which include all the Pango X-associated recombination events (Table 1).

Our analysis of the number of averted mutations reveals a spectrum of plausibility
across all the “robust” recombination events (\ref{fig:averted_mutations}).
Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1).
The “copying pattern” of XA visually confirms
that a high number of recombination-informative sites support the suggested parents (\ref{fig:averted_mutations}).
XW is an example of a marginally well-supported recombinant.
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent.
The copying patterns of all the recombination events are shown in Document S3.

@@

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{fig_recombinant_evidence.pdf}
    \caption{Evidence for recombination in the ARG. All plots show “robust” sc2ts recombination nodes, colored according
to their classification by rebar (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
those identified by \cite{Jackson2021} (see Table \ref{tab:jackson}) are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
    \label{fig:averted_mutations}
\end{figure}

@@

We explored the spatial distribution of the breakpoint locations of
the “robust” recombination events along the genome, and quantified how precisely breakpoints were localized.
For each event, we computed an interval estimate of its breakpoint location,
using its size as a measure of precision (median, 2,207 bases; STAR Method).

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants inferred within each week along the ARG,
focusing on the 356 “robust” events.
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombinant events rose with the global number of cases
(Figure \ref{fig:recombinants_over_time}B; linear regression: $1.0 + 2.7 \; 10^{-7} x$, $p = 1.1 \; 10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure \ref{fig:recombinants_over_time}C; linear regression: $0.9 + 4.8 x$, $p = 3.4\;10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the “expected heterozygosity”).

@@

%%%  All original research papers must include a
%%%  comprehensive and accurate “data and code
%%%  availability” statement within the “resource
%%%  availability” component of the paper before it
%%%  is accepted for publication. These statements
%%%  are structured and consist of three bulleted
%%%  components. Each component must be present.

@@

% The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
% \newline
% (A) Or each panel or group of panels can be described separately, as shown here and below.
% \newline
% (B) Graph of X, Y, and Z.
% \newline
% (C and D) If panels are grouped like this, please explicitly describe each panel, e.g., “Images showing SEM (C) and TEM (D).”
% \newline
% Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}. Example figure created by Cassie Comeau, Cell Press.

@@

Sc2ts (pronounced “scoots”, optionally)
is a method for inferring Ancestral Recombination Graphs (ARGs)
from densely sampled genome sequence data in real time,
in which recombination occurs at a low but significant rate.
The basic idea is to incrementally update an ARG regularly
with the sequences collected since the previous batch \ref{fig:methodology}.
We used daily batches for SARS-CoV-2,
and use days as the time interval for clarity in what follows.
The first step is to find likely “copying paths” under the Li and Stephens (LS) model \cite{Li2003}
for each sequence in the daily batch to the current ARG.
These copying paths will mostly consist of a new sample sequence copying
from a node in the ARG with a small number of mutations,
and often many samples from a daily batch will copy from the same ARG node.
The second step is then to “resolve” these implied polytomies
by using a standard neighbor-joining algorithm \cite{Saitou1987}.
This greedy update strategy inevitably leads to unparsimonious topologies,
and the third step is to then increase the overall parsimony of the inferred ARG
by making some simple topological updates.
Recombination is inferred as an integral and ongoing part of this process,
requiring only a few additional steps to facilitate later analysis.
The result of inference for a given day is then a genealogy recording genetic inheritance
as well as mutation and recombination events
for all of the samples added to the ARG up to that day.

@@

%%% This paragraph needs to be rewritten.
Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997)
to define an alternative formulation of the coalescent with recombination (Hudson, 1983),
the term “Ancestral Recombination Graph” has come to be used
more generally to describe not just realizations of the model,
but any recombinant genetic ancestry (Minichiello and Durbin, 2006).
While there is some subtlety in the details \cite{Wong2024},
we can think of an ARG as being any graph
that encodes the reticulate genetic genealogy of a sample of sequences
under the influence of recombination.

The “succinct tree sequence” (STS) is an ARG data structure
that is both general (in terms of the types of ancestry that can be described)
and computationally efficient \cite{Wong2024}.
Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016},
STS has been extended and applied to
forward-time simulations \cite{Kelleher2018,Haller2019},
calculation of population genetics statistics \cite{Ralph2020} and
ARG inference \cite{Kelleher2019,Wohns2022}.
The STS is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations.
A node represents a particular genome, which may be an observed sample or an inferred genetic ancestor.
The genetic inheritance between a pair of nodes along a segment of genome
is defined by the edge ($l$, $r$, $p$, $c$),
which states that child node $c$ inherited its genome from parent node $p$
from left coordinate $l$ to right coordinate $r$.
A site defines a position on the genome and the ancestral state (allele) at that site.
A mutation records the site and node IDs where a mutation occurs and the derived state (allele).

@@

LS model \cite{Li2003} is an approximation of the coalescent with recombination
that captures many of the key features of the joint processes of mutation and recombination.
It is widely used in a range of applications in population genetics, including ARG inference \cite{Kelleher2019,Gunnarsson2024}.
The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states).
The process of switching between the genomes in the reference panel is governed by a transition matrix,
and mismatches between a focal genome and a genome of the reference panel (from which it is “copied”) are permitted as emissions.

In the original formulation of the LS model (see Appendix S1 of \cite{Li2003}),
the probability of switching between the hidden states (i.e., genomes in a reference panel)
when going from one site to the next is dependent on
(1) the rate of recombination between the sites and
(2) the number of genomes in the reference panel ($n$).
Sc2ts uses the same formulation except that n is set to 1.
This special case of the LS model leads to an intuitive interpretation of the ratio
between the probability of switching and the probability of mismatch:
When considering only two copying paths, the path having k mismatches
and the other path having one switch but no mismatch are equally probable.
We call this parameter k the “recombination penalty”.
This parameter thus allows us to adjust the relative importance of recombination and mutation
when finding the best explanation for a given sample.
For simplicity, we assume that any single base mutation causing a mismatch is equally probable,
ignoring transition-transversion biases and variation of mutation rate along the genome.
As data sampling was dense for SARS-CoV-2, this assumption was considered tolerable.

@@

A Viterbi algorithm can be used to find the most likely “copying path” of a focal sample
through a reference panel of samples under the LS model.
Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS,
which was originally developed to build genome-wide human genealogies
and is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
Using this Viterbi algorithm, sc2ts deterministically infers the copying path of a sample
collected on a certain date through the samples collected on earlier dates in a given ARG.

@@

Although we expected that most samples can be explained by a small number of mutations in the ARG,
we observed two types of outlier samples which had unusually high numbers of mutations:
(1) “time travelers” and (2) “saltation” lineages.
Time traveling samples are caused by errors in their collection dates.
For example, if a sample is actually collected 10 months after its reported collection date,
then 20 mutations (instead of, say, 0 or 1) are needed to explain it
using the ARG available on its collection date,
assuming a substitution rate of 2 events per genome per month.
Samples from “saltation” lineages (particularly the major VOCs) are characterized
by high numbers of mutations by the time of their detection.
Both types of outlier samples are poorly fit to the ARG when they appear on their collection dates.
To handle these outlier samples, we introduce three mechanisms:
(1) HMM cost-based rejection,
(2) retrospective sample grouping, and
(3) unconditional seeding.

@@

\subsection{Analysis of Pango lineage ``origination'' events}

>>>>

UShER is a maximum parsimony method developed during the pandemic to infer phylogenies
from large numbers of densely sampled SARS-CoV-2 genome sequences \cite{Turakhia2021}.
UShER has generated phylogenies updated daily \cite{McBroome2021} and
has proven to be a powerful resource that successfully enabled the world
to keep up with an evolving disease (e.g., through the Nextstrain web portal \cite{Hadfield2018}).
These phylogenies are used to identify and monitor
potentially epidemiologically important ``Pango'' lineages \cite{Rambaut2020,McBroome2024},
which might have acquired features to become Variants of Concern (VOC),
for example, increased transmissibility, disease severity, and immune escape.
Beyond supporting real-time genomic epidemiology,
UShER phylogenies have been used in a multitude of evolutionary and epidemiological studies of SARS-CoV-2
(e.g., \cite{Karthikeyan2022,Obermeyer2022,Turakhia2022,Bloom2023}).
UShER uses a tree to represent evolutionary change via mutation, assuming no recombination \cite{Turakhia2021}.
However, given extensive evidence for recombination in SARS-CoV-2 \cite{Jackson2021,VanInsberghe2021,Focosi2022,Gutierrez2022,Ignatieva2022,Tamura2023},
the tree-like representation of UShER is no longer tenable.

@@

Here, we present
(1) sc2ts, a new method to infer ARGs in real time by building genealogies forward in time, and
(2) an ARG for SARS-CoV-2 with ~2.48 million genome sequences.
We used the ``Viridian'' dataset, which is purged of various systematic errors and artifacts \cite{Hunt2024} and
show that this ARG is highly congruent to an UShER phylogeny built from the same dataset.
Compared to previously identified recombinants,
we show that this ARG contains the vast majority of epidemiologically relevant signals of recombination,
while adding significant nuance and texture.
This ARG – which captures the major signals of evolution
through phylogenetic relationships, point mutations, deletions, and recombination –
is a resource of unprecedented detail and precision, and
should provide a strong basis for future studies of SARS-CoV-2.

@@

Sc2ts uses collection dates to build a time-resolved ARG.
When examining preliminary ARGs, we found samples that introduced large numbers of mutations (and reversions),
likely due to errors in their collection dates
(for example, a large batch of samples dated 31 December, 2020 contains samples from Omicron BA.1,
which is estimated to have emerged in November, 2021; REF).
To preclude adding such ``time traveling'' samples, we filtered samples by ``HMM cost''.
We define the ``HMM cost'' of a sample to be
the total penalty cost for switches and mismatches (mutations) in its copying path.
We parameterize the LS model such that each mismatch has a penalty cost of one
and a switch has a penalty cost of $k$ (here, set to 4),
which determines how much mutation is favored over recombination when accounting for a 
sample (Figure~\ref{fig:methodology}).
If the HMM cost of a sample exceeds a threshold (here, set to 7),
sc2ts does not immediately add the sample to the ARG.

@@

Input genome alignments are masked in sc2ts
so that deletions and ambiguous bases are treated as missing data by the HMM,
and once the samples are incorporated into the ARG,
these bases are imputed to carry the values inherited by the node in question.
Deletions are an important part of SARS-CoV-2 evolution \cite{Jeronimo2023}, however,
and to incorporate them into the ARG we identified a set of [XXX] sites
most frequently involved in deletions [cite source?]
and post-hoc mapped alignment data for these sites to the ARG using parsimony.
Although sites are treated independently in this approach,
the approximation successfully identified the most frequent deletions,
such as the recurrent 21991-21993 (See Document~\ref{sec:supplement_dels}
for additional analysis).
The node representing the origin of Alpha, with 290,407 descendants,
is correctly associated with key lineage-defining deletions:
11288-11296 (ORF1A del3675-3677),
21765-21770 (Spike del69-70),
21991-21993 (Spike del144), and
28271 (non-coding).
The Spike 69-70 deletion, which causes S-Gene Target Failure \cite{Walker2021},
also emerges twice in the origin of Omicron -
in the node leading to the BA.1 variant (342,099 descendants),
and the node leading to the BA.4 variant (189,746 descendants).
To speed up inference and to avoid issues caused by sites enriched for errors [citation?],
sc2ts masks out a set of 100 sites identified as the most highly homoplasic
by running on a subset of the data (STAR Methods).
Of these sites, 9 are in the set of [1XX] ``problematic sites'' 
masked by UShER \cite{Hunt2024}.
In addition to remapping the data for sites identified as being involved in deletions,
we added mutations for XX of these highly recurrent sites
that are used in Pango designation \cite{Rambaut2020} by post-hoc parsimony.
Each node in the ARG was then assigned to one of 2026 Pango lineages using pangolin (STAR Methods).
All but 8293 sampled sequences (0.33\%) agreed with the source sequences in terms of Pango lineage assignment
(comparable to the 0.27\% disagreement between pangolin-data 
versions 1.21 and 1.29 in the source Viridian metadata).
A majority of these lineages (1469 of 2058) have monophyletic origins within the ARG,
increasing to 1779 (86\%) when we allow for multiple sibling origination 
nodes (Document~\ref{sec:supplement_pango_lineages}).

@@

As new data became available during the pandemic,
volunteers continually detected recombinants which might have concerning features,
particularly increased transmissibility, disease severity, and
ability to escape natural or vaccine-induced immunity (e.g., XBB\cite{Tamura2023}).
The process of gathering evidence for and characterizing such recombinants is manual,
involving visual inspection of phylogenies for unusual branches
caused by forcibly placing recombinant sequences onto them,
typically followed by visual analysis of genome sequence alignments
to identify parental lineages and breakpoint locations.
This process has resulted in the designation of recombinant Pango lineages
(whose aliases begin with an ``X''), to which we refer as ``Pango X''.
Below, we summarize our analysis of the Pango Xs in the ARG
(for extended analyses, see Document S1 and Document S2).

We find strong evidence for the recombination events leading to 17 Pango Xs in the ARG (Table 1),
which passed the detection-level sc2ts quality control filters (Figure S1).
These events are supported by
high numbers of averted mutations (Table 1) and
high numbers of loci matching their suggested parents (see their copying patterns in Figure SX).
Among these Pango Xs are
XA, which is the first designated Pango X that showed signs of community transmission \cite{Jackson2021}, and
XBB, which escaped vaccine-induced immunity and gained global prevalence \cite{Tamura2023}.
The recombinant origin of XA is well explained in the ARG by two parents and one mutation (\ref{fig:subgraphs}; Table 1),
with sc2ts correctly identifying its parental Pango lineages and breakpoint location.

Furthermore, some of the Pango Xs above originate from shared recombination events in the 
ARG (Table 1).
An example of a set of ``nested'' closely related Pango Xs consists of XZ, XAC, XAD, XAE, and XAP,
which descend from a single recombination event rather than five independent events in the ARG (\ref{fig:subgraphs}).
These Pango Xs were designated around the same time.
These results highlight that the ARG can help clarify the relationships between recombinants,
allowing more precise interpretation of evidence for recombination.

@@

\begin{figure}[h]
\centering
\includegraphics[width=0.85\linewidth]{pango_x_subgraphs.png}
\caption{Illustrative subgraphs in the ARG. Subgraphs around the XA-associated recombination node (A); the recombination node associated with a group of ``nested'' Pango Xs (B); and the MRCA of the non-recombinants XN and XAU (C). For more detailed visualizations of these subgraphs, see Document S2.}
\label{fig:subgraphs}
\end{figure}

@@

We also compared the sensitivity of sc2ts and rebar.
Of all the ``robust'' recombination events detected by sc2ts,
167 (43\%) were also detected as recombinants by rebar (STAR Methods).
We hypothesize that this disagreement arose from differences in sensitivity between these methods.
Consistent with this, we find that rebar needs a higher level of divergence between parents
to correctly classify a sequence as recombinant, as shown by
a shorter time to their MRCA (\ref{fig:averted_mutations}) and
lower distance between their Pango identity (\ref{fig:averted_mutations}).

@@

Sc2ts provides a method to facilitate the discovery of recombination events,
by comparing how well a sample fits the ARG under a model allowing recombination
versus a model disallowing recombination, as measured by the number of resulting mutations (\ref{fig:methodology}).
The more mutations are ``averted'' by allowing recombination, the stronger the evidence for recombination.
We calculated the number of averted mutations for all the ``robust'' recombination events (STAR Methods),
which include all the Pango X-associated recombination events (Table 1).

Our analysis of the number of averted mutations reveals a spectrum of plausibility
across all the ``robust'' recombination events (\ref{fig:averted_mutations}).
Examples of highly plausible recombinants are XA and XBB (for an extended analysis, see Document S1).
The ``copying pattern'' of XA visually confirms
that a high number of recombination-informative sites support the suggested parents (\ref{fig:averted_mutations}).
XW is an example of a marginally well-supported recombinant.
The initial evidence for proposing XW for designation is four nucleotides associated with the suggested BA.1 parent.
The copying patterns of all the recombination events are shown in Document S3.

@@

%%% TODO: Old version as placeholder. This figure is not yet finalised.
\begin{figure}[h]
    \centering
    \includegraphics[width=1\linewidth]{fig_recombinant_evidence.pdf}
    \caption{Evidence for recombination in the ARG. All plots show ``robust'' sc2ts recombination nodes, colored according
to their classification by rebar (recombinant: purple, non-recombinant: gray).
(A) Upper: jittered scatterplot comparing number of mutations when recombination allowed ($y$: right axis)
versus an alternative copying path disallowing recombination ($x$: top axis);
recombination nodes in Table \ref{tab:pango_x_lineages} are labelled starting with `X' and
those identified by \cite{Jackson2021} (see Table \ref{tab:jackson}) are prefixed with a `J', e.g.
Jackson Group B (JGB) and Jackson Singleton recombinant 1 (JS1). 
Point sizes reflect number of descendants.
Lower: recombination nodes classified by number of mutations averted by allowing recombination (i.e. $x$-$y$).
(B) Distribution of time to the MRCA of the parents of each robust recombination node.
(C) Distribution of the Pango lineage distance between parents.
(D) Copying patterns leading to XA, XBB, XBB.1, and the recombinant origin of XZ+XAC+XAD+XAE+XAP;
each recombination node is a mix of variants matching the left parent (green), the right parent (blue), and de-novo mutations (gold).}
    \label{fig:averted_mutations}
\end{figure}

@@

We explored the spatial distribution of the breakpoint locations of
the ``robust'' recombination events along the genome, and quantified how precisely breakpoints were localized.
For each event, we computed an interval estimate of its breakpoint location,
using its size as a measure of precision (median, 2,207 bases; STAR Method).

@@

We used the COVID-19 Data Repository at Johns Hopkins University \cite{Dong2020}
to relate the timing of recombination events with global case numbers.
Case numbers ($n$) were binned by week from January, 2020 to March, 2023 and
sub-divided by the proportion in each Scorpio label inferred within the ARG (shading in \ref{fig:recombinants_over_time}).
These lineages correspond loosely to a group of related lineages within a major VOC in the Viridian dataset.
Alongside the case numbers,
the black curve shows the number of recombinants inferred within each week along the ARG,
focusing on the 356 ``robust'' events.
Recombination events were inferred more often during the Delta and Omicron waves, when case counts were high.
Overall, the number of recombinant events rose with the global number of cases
(Figure \ref{fig:recombinants_over_time}B; linear regression: $1.0 + 2.7 \; 10^{-7} x$, $p = 1.1 \; 10^{-7}$).
Recombination events also rose with the diversity of lineages in each week
(Figure \ref{fig:recombinants_over_time}C; linear regression: $0.9 + 4.8 x$, $p = 3.4\;10^{-7}$),
measured as the chance that two randomly drawn viruses have different Scorpio designations
($H$, referred to as the ``expected heterozygosity'').

@@

%%%  All original research papers must include a
%%%  comprehensive and accurate ``data and code
%%%  availability'' statement within the ``resource
%%%  availability'' component of the paper before it
%%%  is accepted for publication. These statements
%%%  are structured and consist of three bulleted
%%%  components. Each component must be present.

@@

% The figure legend can be all one paragraph and describe the images (A), graphs (B), and plots (C), etc., together.
% \newline
% (A) Or each panel or group of panels can be described separately, as shown here and below.
% \newline
% (B) Graph of X, Y, and Z.
% \newline
% (C and D) If panels are grouped like this, please explicitly describe each panel, e.g., ``Images showing SEM (C) and TEM (D).''
% \newline
% Please define all scale and error bars, and please review the Cell Press figure guidelines before submission: \href{https://www.cell.com/figureguidelines}{https://www.cell.com/figureguidelines}. Example figure created by Cassie Comeau, Cell Press.

@@

Sc2ts (pronounced ``scoots'', optionally)
is a method for inferring Ancestral Recombination Graphs (ARGs)
from densely sampled genome sequence data in real time,
in which recombination occurs at a low but significant rate.
The basic idea is to incrementally update an ARG regularly
with the sequences collected since the previous batch \ref{fig:methodology}.
We used daily batches for SARS-CoV-2,
and use days as the time interval for clarity in what follows.
The first step is to find likely ``copying paths'' under the Li and Stephens (LS) model \cite{Li2003}
for each sequence in the daily batch to the current ARG.
These copying paths will mostly consist of a new sample sequence copying
from a node in the ARG with a small number of mutations,
and often many samples from a daily batch will copy from the same ARG node.
The second step is then to ``resolve'' these implied polytomies
by using a standard neighbor-joining algorithm \cite{Saitou1987}.
This greedy update strategy inevitably leads to unparsimonious topologies,
and the third step is to then increase the overall parsimony of the inferred ARG
by making some simple topological updates.
Recombination is inferred as an integral and ongoing part of this process,
requiring only a few additional steps to facilitate later analysis.
The result of inference for a given day is then a genealogy recording genetic inheritance
as well as mutation and recombination events
for all of the samples added to the ARG up to that day.

@@

%%% This paragraph needs to be rewritten.
Originally introduced (Griffiths, 1991; Griffiths and Marjoram, 1997)
to define an alternative formulation of the coalescent with recombination (Hudson, 1983),
the term ``Ancestral Recombination Graph'' has come to be used
more generally to describe not just realizations of the model,
but any recombinant genetic ancestry (Minichiello and Durbin, 2006).
While there is some subtlety in the details \cite{Wong2024},
we can think of an ARG as being any graph
that encodes the reticulate genetic genealogy of a sample of sequences
under the influence of recombination.

The ``succinct tree sequence'' (STS) is an ARG data structure
that is both general (in terms of the types of ancestry that can be described)
and computationally efficient \cite{Wong2024}.
Initially developed to facilitate large-scale coalescent simulations \cite{Kelleher2016},
STS has been extended and applied to
forward-time simulations \cite{Kelleher2018,Haller2019},
calculation of population genetics statistics \cite{Ralph2020} and
ARG inference \cite{Kelleher2019,Wohns2022}.
The STS is based on a simple tabular representation,
which defines a set of nodes, edges, sites and mutations.
A node represents a particular genome, which may be an observed sample or an inferred genetic ancestor.
The genetic inheritance between a pair of nodes along a segment of genome
is defined by the edge ($l$, $r$, $p$, $c$),
which states that child node $c$ inherited its genome from parent node $p$
from left coordinate $l$ to right coordinate $r$.
A site defines a position on the genome and the ancestral state (allele) at that site.
A mutation records the site and node IDs where a mutation occurs and the derived state (allele).

@@

LS model \cite{Li2003} is an approximation of the coalescent with recombination
that captures many of the key features of the joint processes of mutation and recombination.
It is widely used in a range of applications in population genetics, including ARG inference \cite{Kelleher2019,Gunnarsson2024}.
The LS model is an HMM that models a focal genome as a sequence of nucleotides (i.e. the observed states),
which are probabilistically emitted as an imperfect mosaic of a set of genomes in a reference panel (i.e., the hidden states).
The process of switching between the genomes in the reference panel is governed by a transition matrix,
and mismatches between a focal genome and a genome of the reference panel (from which it is ``copied'') are permitted as emissions.

In the original formulation of the LS model (see Appendix S1 of \cite{Li2003}),
the probability of switching between the hidden states (i.e., genomes in a reference panel)
when going from one site to the next is dependent on
(1) the rate of recombination between the sites and
(2) the number of genomes in the reference panel ($n$).
Sc2ts uses the same formulation except that n is set to 1.
This special case of the LS model leads to an intuitive interpretation of the ratio
between the probability of switching and the probability of mismatch:
When considering only two copying paths, the path having k mismatches
and the other path having one switch but no mismatch are equally probable.
We call this parameter k the ``recombination penalty''.
This parameter thus allows us to adjust the relative importance of recombination and mutation
when finding the best explanation for a given sample.
For simplicity, we assume that any single base mutation causing a mismatch is equally probable,
ignoring transition-transversion biases and variation of mutation rate along the genome.
As data sampling was dense for SARS-CoV-2, this assumption was considered tolerable.

@@

A Viterbi algorithm can be used to find the most likely ``copying path'' of a focal sample
through a reference panel of samples under the LS model.
Specifically, sc2ts uses a Viterbi algorithm modified to operate on STS,
which was originally developed to build genome-wide human genealogies
and is capable of scaling to biobank-scale datasets \cite{Kelleher2019}.
Using this Viterbi algorithm, sc2ts deterministically infers the copying path of a sample
collected on a certain date through the samples collected on earlier dates in a given ARG.

@@

Although we expected that most samples can be explained by a small number of mutations in the ARG,
we observed two types of outlier samples which had unusually high numbers of mutations:
(1) ``time travelers'' and (2) ``saltation'' lineages.
Time traveling samples are caused by errors in their collection dates.
For example, if a sample is actually collected 10 months after its reported collection date,
then 20 mutations (instead of, say, 0 or 1) are needed to explain it
using the ARG available on its collection date,
assuming a substitution rate of 2 events per genome per month.
Samples from ``saltation'' lineages (particularly the major VOCs) are characterized
by high numbers of mutations by the time of their detection.
Both types of outlier samples are poorly fit to the ARG when they appear on their collection dates.
To handle these outlier samples, we introduce three mechanisms:
(1) HMM cost-based rejection,
(2) retrospective sample grouping, and
(3) unconditional seeding.

@@

\subsection{Analysis of Pango lineage ``origination'' events}
\label{sec:supplement_pango_lineages}
To analyze how well the ARG reflects the phylogenetic structure implicit in the
Pango lineage naming system, we considered the Pango assignments generated by
Pangolin on the alignments for each node in the ARG (STAR Methods). If the ARG
perfectly reflected the Pango lineage structure, each lineage would form a
clade descending from a single originating node with that label. We identified
putative origination nodes among those labeled with a given lineage as the node
with the maximum number of descendant samples and the earliest time (if there
is a tie).
